\chapter{Proving Full Abstraction}
\label{chap:InformalProof}
Some formal techniques of proving the correctness of a compilation scheme offering full abstraction exist and the idea of one of these proof techniques, based on trace semantics, is sketched in \myref{sec}{sec:prooftechniques}.
%Using the ideas of this proof technique, \myref{sec}{sec:informalproof} aims to give a short and informal proof for the compilation scheme presented earlier.

\section{Formal Proof Techniques}
\label{sec:prooftechniques}
Recalling from \myref{sec}{sec:fullabstraction}, \emph{full abstraction} is a compiler property.
It states that \emph{contextual equivalence} for source-level objects is preserved by and reflected from their target-level translations.
\[
    O_{1} \simeq O_{2} \iff \compiled{O_{1}} \simeq \compiled{O_{2}}
\]

Proving full abstraction of a compilation scheme, i.e. proving that it is secure, requires a proof for soundness and completeness.
\begin{description}
\item[Soundness]
Soundness expresses that the compilation of two source-level objects does not `introduce' contextual equivalence in the target language.
Instead, for the target-level objects to be contextually equivalent, the source-level objects have to contextually equivalent already.
\[
    \compiled{O_{1}} \simeq \compiled{O_{2}} \implies O_{1} \simeq O_{2}
\]
Soundness corresponds closely to the informal notion of compiler \emph{correctness}.
Indeed, formulating the logically equivalent contrapositive of the soundness property gives:
\[
    O_{1} \not\simeq O_{2} \implies \compiled{O_{1}} \not\simeq \compiled{O_{2}}
\]

This expresses that two contextually \emph{un}equivalent source-level objects result in contextually \emph{un}equivalent translations.
If a compilation scheme would \emph{not} be sound, there would exist two contextually \emph{un}equivalent source-level objects, whose translations would be contextually equivalent.

Clearly, such a compiler does not function `correctly', as there is a context in which the source-level objects behave differently, but the translations do not.
One of these translations does not accurately behave like the source-object it is derived from.
\item[Completeness]
Completeness says that all contextually equivalent source-level objects are translated to contextually equivalent target-level objects.
It expresses that the contextual equivalence of source-level objects, which provides certain security guarantees, are preserved when compiling.
\[
    O_{1} \simeq O_{2} \implies \compiled{O_{1}} \simeq \compiled{O_{2}}
\]
\end{description}

As most compilers are expected to be `correct' or sound, the most important part of the full abstraction proof is the proof of completeness.
Proving completeness can be done using \emph{trace semantics} and looking at the contrapositive of completeness.
\[
    \compiled{O_{1}} \not\simeq \compiled{O_{2}} \implies O_{1} \not\simeq O_{2}
\]
The following section will detail how trace semantics can be used to prove completeness of a compilation scheme.

\subsection{Trace Semantics}
\label{sec:tracesemantics}
Trace semantics~\cite{Rathke, Patrignani:TraceSemantics} for the low-level language describe the behavior of a program $\compiled{P}$ within a context $\compiled{O_C}$ as all interactions that happen between the context $\compiled{O_C}$ and the program $\compiled{P}$.
Only the interaction between context $\compiled{O_C}$ and program $\compiled{P}$ is described: this means that trace semantics do \emph{not} capture any \emph{internal} operation.
The interactions described correspond with any exchange of data or information between context $\compiled{O_C}$ and program $\compiled{P}$.

Trace semantics produce a trace for a program $\compiled{P}$ executing within a context $\compiled{O_C}$.
A trace is a sequence of \emph{labels} that correspond with the executed instructions.
In order to restrict the description to the \emph{interaction} between program $\compiled{P}$ and context $\compiled{O_C}$, not all instructions are given a label.
Instead, only the \lsttext{call} and \lsttext{ret} instruction are labeled.
Trace semantics track whether the instruction was executed by program $\compiled{P}$ or by context $\compiled{O_C}$.
A possible syntax~\cite{Patrignani:TraceSemantics} for traces could be the one given by \myref{fig}{fig:tracesemanticssyntax}.
\newcommand{\fs}{\;\;\;}
\vspace{-1.5em}
{\setlength{\abovecaptionskip}{0pt}
\begin{figure}[H]
\begin{align*}
\begin{aligned}
\text{Trace\ } T\ ::=\ L\ |\ T\cdot L \fs & \fs L\ ::=\ a\ |\ \tau \fs & \fs a\ ::=\ g?\ |\ g! \fs & \fs g \::=\ \kw{call}\ p(\overline{v})\ |\ \kw{ret}\ v
\end{aligned}
\end{align*}
\caption{Syntax for trace semantics\label{fig:tracesemanticssyntax}.}
\end{figure}
}
\vspace{-1em}
In this syntax, labels are defined to be actions $a$ or $\tau$, where an action $a$ is observable, and $\tau$ is not.
It uses $!$ and $?$ to track whether the action was performed in program $\compiled{P}$ or context $\compiled{O_C}$, respectively.
In a \lsttext{call}, $p$ represents the address that was called and $\overline{v}$ the values passed in registers.
In a \lsttext{ret}, $v$ represents the value in the return register.

Full abstraction proofs based on trace semantics, for example the ones given by Patrignani et al.~\cite{Patrignani,Patrignani:TraceSemantics} or Agten et al.~\cite{Agten:2012:SCM:2354412.2355247}, first need to show that the operational semantics of the target language are fully abstract to the proposed low-level trace semantics.

This is only the case if the proposed trace semantics in fact do capture \emph{all} exchange of information and interactions between context $\compiled{O_C}$ and program $\compiled{P}$.
For the trace semantics with syntax given by \myref{fig}{fig:tracesemanticssyntax}, this is only true if the operational semantics prohibit the exchange of information between context $\compiled{O_C}$ and program $\compiled{P}$ through memory or registers other than the return value register.

If the operational semantics do \emph{not} prohibit the exchange of information through memory or registers other than the return register, then the syntax of labels must be changed to a syntax that captures these channels of communication, as stated by Curien~\cite{Curien2007301} and worked out by Patrignani et al.~\cite{Patrignani:TraceSemantics}.
\\[1em]
If the trace semantics are fully abstract w.r.t the operational semantics, then the proof of full abstraction of the compilation scheme becomes easier.
As traces then fully capture all the interaction of context object $\compiled{O_{c}}$ with a target-level object $\compiled{O_{1}}$, two target-level objects $\compiled{O_1}$ and $\compiled{O_2}$ are indistinguishable with respect to a context if and only if their traces $T_1$ and $T_2$ are the same.
It follows that the existence of a context $\compiled{O_C}$ that can distinguish between $\compiled{O_1}$ and $\compiled{O_2}$, implies that they produce different traces $T_1$ and $T_2$ for this context $O_C$.

As a consequence, the contrapositive statement of completeness can be rewritten formally as:
\[
     \text{Traces}_L(\compiled{O_1}) \neq  \text{Traces}_L(\compiled{O_2}) \implies O_{1} \not\simeq O_{2}
\]

Proving this statement is simpler than proving the unmodified contrapositive.
A proof for this statement would for example consist of an algorithm that can create a high-level context $O_C$ that is able to distinguish $O_1$ from $O_2$ using their differing low-level traces $T_1$ and $T_2$.
This context $O_C$, called the witness~\cite{Patrignani}, shows that $O_1$ and $O_2$ aren't contextually equivalent on the high-level.

If an algorithm exists that for any such pair of traces can construe a high-level context $O_C$, then the existence of this algorithm proves full abstraction of the compilation scheme.

\section{Conclusion}
A proof of full abstraction of the compilation scheme can be achieved by proving a low-level trace semantics exists that is fully abstract with the operational semantics of the low-level language.
The trace semantics must capture \emph{all exchange of information} or \emph{interactions} between the low-level program and its context.
Curien~\cite{Curien2007301} stated that one can make sure the labels of trace semantics capture every interaction between program and context by either:
\begin{itemize}
\item making the labels more expressive, so that they can capture the additional channels of communication that can be used, besides the registers at calls and the return register at returns;
\item modifying the operational semantics, so that they restrict the communication between context and program to those channels captured by the labels.
\end{itemize}

In this work, as well as that of Agten et al.~\cite{Agten:2012:SCM:2354412.2355247} and Patrignani et al.~\cite{Patrignani}, the compiler modifies the operational semantics so that it indeed limits the exchange of information in this way.




%Trace semantics~\cite{Rathke, Patrignani:TraceSemantics} describe the behavior of a component as a series of method calls and returns.

%Full abstraction proofs based on trace semantics, such as~\cite{Patrignani,Agten:2012:SCM:2354412.2355247}, first show that the operational semantics of the target language are equivalent to the proposed trace semantics.
%As a consequence, the interaction of context objects $O_{c}$ with target-level objects $O_{1}$ and $O_{2}$ in the definition of contextual equivalence can be represented by traces $T_{1}$ \& $T_{2}$.

%If two target-level objects $\compiled{O_1}$ and $\compiled{O_2}$ are not contextually equivalent, their traces $T_1$ and $T_2$ must be different.
%The proof of fully abstract compilation then must show that if two source-level objects $O_1$,$O_2$ have different target-level traces $T_1$ and $T_2$, a source-level context object $O_C$ can be created from those two traces.
%This $O_C$ will be able to serve as a source-level object that differentiates between two objects $O_1$ and $O_2$.
%The existence of such an object effectively means the source-level objects $O_1$ and $O_2$ are not contextually equivalent, which concludes the completeness proof.



%\section{Informal Proof}
%\label{sec:informalproof}
%
%This section gives a short and informal reasoning that suggests that the compilation scheme presented in \myref{sec}{sec:formalizedcompiler2} is fully abstract.
%The argument is highly influenced by the formal proof technique described in \myref{sec}{sec:tracesemantics}.
%\\[1em]
%The first important step in a formal proof would be showing that the operational semantics of the target language have an equivalent trace semantics.
%The secure compilation ensures this by limiting the information passed between insecure context and secure code.
%
%This is achieved by clearing flags and registers~\cite{Agten:2012:SCM:2354412.2355247} when control flow passes from secure code to insecure context.
%Secondly, the leaking of information about memory locations is prevented.
%This is partly achieved by the use of masking~\cite{Patrignani}, and for the other part by making fields available using the generic structure value entry point and a (masked) reference to a frame.
%
%Information leakage about the order of definitions, whether it be value definitions of structure definitions, is prevented by imposing an alphabetical ordering on the definitions.
%An alphabetic reordering at compilation makes sure that the order of fields within a frame, or frames within the f-list is insensitive to a reordering of the source-level code.
%The addition of functors requires additional care to ensure that there is no way of telling whether a structure was created by a static binding or using functor application.
%\\[1em]
%The next step would be showing that any pair of low-level traces $T_1$ and $T_2$ that differentiate two compiled objects $\compiled{O_1}$ and $\compiled{O_2}$ makes it possible to construct a high level context $O_c$ that could differentiate between $O_1$ and $O_2$.
%To ensure this, tampering with the control flow must be prevented.
%
%In the compilation scheme presented here, the control flow is secured by the memory access model, which limits the execution rights of the insecure context to the entry points specified.
%Only by jumping to these entry points can the execution switch between the insecure context and the secure code.
%
%However, jumping to one of the entry points corresponds to making a call to something in the API of the source-level object.
%Because every call from insecure context to secure code listed in the trace corresponds to a call to the API of the source-level object, it can be shown that a pair of traces can be merged to a source-level context object. 
%This source-level context object will be able to differentiate between the source-level objects, just as the pair of traces could for the target-level translations.
%\\[1em]
%The existence of such a source-level context object shows that the source-level objects are not contextually equivalent.
%This proves the completeness property, using its contrapositive form $\compiled{O_{1}} \not\simeq \compiled{O_{2}} \implies O_{1} \not\simeq O_{2}$.
%%Find other references as well!