\begin{tabularx}{\textwidth}{@{}l c X@{}}
\compile{$\overline{me}$} \makes \compile{sort($\overline{me}$)}\\

\intertextt{
Then, all module expressions can be translated individually.
The signature \cmath{SigId} used in the ascription of \cmath{StrId} is saved:
}

\compile{\cmath{me, \overline{me}}} \makes \compile{sort($\overline{me}$)} \\
\compile{\cmath{StrId:SigId = \overline{d}}} \makes \compile{$\overline{d}$}\cmath{^{SigId}}\\

\intertextt{
All fields in a module are sorted alphabetically as well. 
This is done for the same reason as the sorting of module expressions: two \MiniML\ programs \cmath{P} differing only in the order of their fields are contextually equivalent.
To prevent leakage of this ordering in the target-level, outputting the fields in the target language happens in a fixed ordering.
}

\compile{$\overline{d}$}\cmath{^{SigId}} \makes \compile{sort($\overline{d}$)}\cmath{^{SigId}} \\

\intertextt{
The compiler can now continue by translating the elements specified within the module expression.
Every definition is translated separately.
}
\compile{$\overline{d}$}$^{\mathit{SigId}}$ \makes \compile{$d$}$^{\mathit{SigId}}$;\compile{$\overline{d}$}$^{\mathit{SigId}}$\\

\intertextt{
The translation of a definition depends on a number of things.
\begin{itemize}
\item Whether the definition is that of a type or that of a field.
%The definition of a type is not translated to llvm.
\item Whether or not the value is local to the structure \cmath{StrId}.
This depends on the value being declared in the signature \cmath{SigId}.
\end{itemize}
}

\compile{\lsttext{type}$\ \ova\ t= \tau$} \makes \lsttext{\%t = type $\ \{\tau\}$} \nl
\annot{The type and its definition are tracked.
The type is assigned a unique integer $\ \tau_{int}$ to identify it.
This way, when a masked pointer is received in an entry stub, its type can be checked.}\\
& & \\
\compile{\lsttext{val} $id = e : \tau$}$^{\mathit{SigId}}$ \makes \lsttext{define private\ $\tau$* @}$id$\lsttext{_internal()\{} \compile{$e$}$^{\mathit{SigId}}$ \lsttext{\}} \nl
\annot{The next part is only included if the value is declared in the signature}\nl
\lsttext{define \%int @}$id$ \lsttext{()\{} \nl
\annot{\setlength{\leftskip}{3.5ex} Switch stack, move parameters, add entry point in spm for @\cmath{id}}\nl
\longspace \lsttext{\%0 = call $\ \tau$* @}$id$\lsttext{_internal()} \nl
\longspace \lsttext{\%1 = ptrtoint $\ \tau$* \%0 to \%int} \nl
\longspace \lsttext{\%2 = call \%int @mask(\%int \%1,\%int $\ \tau_{int}$)} \nl
\longspace [Switch stack, clear registers] \nl
\longspace \lsttext{ret \%int \%2} \nl
\lsttext{\}} \\
& & \\
\compile{\lsttext{fun\ $\cmath{id\ \overline{x} = e : \tau}$}}$^{\mathit{SigId}}$ \makes \lsttext{define private\ $\tau$* @}$id$\lsttext{_internal($\mathit{\overline{\tau_{x}*}\ \overline{x}}$)\{}\nl
\longspace \compile{$e$}$^{\mathit{SigId}}$ \nl
\lsttext{\}} \nl
\annot{The next part is only included if the value is declared in the signature}\nl
\lsttext{define \%int @}$id$ \lsttext{($\mathit{\overline{\%int}\ \overline{y}}$)\{} \nl
%\longspace [Switch stack, move parameters, add entry point in spm for @\cmath{id}]\nl
\annot{\setlength{\leftskip}{3.5ex} Switch stack, move parameters, add entry point in spm for @\cmath{id}}\nl
\longspace \lsttext{$\overline{\%z}$ = call \%int @unmask($\overline{\%int}\ \overline{\%y}$)} \nl
\longspace \lsttext{$\overline{\%t}$ = call \%int @unmasktype($\overline{\%int}\ \overline{\%y}$)} \nl
\longspace \lsttext{$\overline{\%c}$ = icmp eq \%int $\ \overline{\%t}$\ $\overline{\%\tau_{x,int}}$}\nl
\longspace \lsttext{br i1 $\ \overline{\%c}$, label \%Continue, label \%Error}\nl
\nl
\longspace \lsttext{Continue:} \nl
\longspace \lsttext{$\overline{\%x}$ = inttoptr \%int\ $\overline{\%z}$ to $\ \overline{\tau_{x}*}$} \nl
\longspace \lsttext{\%0 = call $\ \tau$* @}$id$\lsttext{_internal($\overline{\tau_{x}*}\ \overline{\%x}$)} \nl
\longspace \lsttext{\%1 = ptrtoint $\ \tau$* \%0 to \%int} \nl
\longspace \lsttext{\%2 = call \%int @mask(\%int \%1, \%int $\ \tau_{int}$)} \nl
\longspace [Switch stack, clear registers and flags] \nl
\longspace \lsttext{ret \%int \%2} \nl
\nl
\longspace \lsttext{Error:}\nl
\longspace \lsttext{call void @exit(i32 -1)}\nl
\longspace \lsttext{unreachable}\nl
\lsttext{\}} \\ 
\intertextt{The code of the \lsttext{@mask}, \lsttext{@unmask} and \lsttext{@unmasktype} functions is given in \myref{lst}{lst:maskcode} in \myref{app}{app:LLVMCode}. 
It provides a naive implementation of \lsttext{@mask}, \lsttext{@unmask} and \lsttext{@unmasktype} functions using linked lists.}
\intertextt{If a function receives polymorphic arguments, for example \lsttext{createPair :: a -> a -> Pair a}, the internal functions arguments are of type \lsttext{\%tyvar*}.
When the function is called from insecure code, the stub loads the parameters from the masking list and wraps them in a \lsttext{\%tyvar}.

Before passing these arguments to the internal function, the compiler inserts code to check the consistency of the underlying type equation, using the \lsttext{@tyvarcheck} function.
The code implementing \lsttext{@tyvarcheck} is given in \myref{lst}{lst:tyvarcheck} in \myref{app}{app:LLVMCode}.

As an example, \myref{sec}{sec:PolymorphicExample} in \myref{app}{app:LLVMCode} provides an example of the compilation of a polymorphic function.
}
\end{tabularx}