\chapter{Conclusion}
\label{chap:conclusion}
First, this chapter discusses related work in \myref{sec}{sec:relatedwork}.
It continues by discussing possible future work in section \myref{sec}{sec:FutureWork}.
A conclusion to the work is formulated as well in \myref{sec}{sec:Conclusion}

\section{Related Work}
\label{sec:relatedwork}
Extensive work trying to preserve the security of a source languages when compiling exists.
The idea of using full abstraction to formalize secure compilation is introduced by Abadi~\cite{Abadi}.

Different techniques to preserve security even in the low-level computing model were developed, for example the use of \emph{Adress Space Layout Randomization} or ASLR to \emph{prevent} reliable memory manipulation by an attacker.
This technique randomly changes the arrangement of important program portions.
For example, the location of the stack, heap and base of the executable might change location within a process's address space.
This reduces the reliability of an attacker's view of the process's address space, making it harder if not impossible for attackers to reliably jump to certain exploitable parts of the code, or to reliably overwrite a certain value in memory.

% introduced by Abadi~\cite{AbadiASLR} as well.
The idea of ASLR catched on, and ASLR saw implementations in common operating systems such as Windows Vista, OS X Mountain Lion and some Linux distributions. 
The idea also raised scientific study, for example by Abadi and Plotkin~\cite{AbadiASLR} or Jagadeesan, et al.~\cite{Jagadeesan} and criticism~\cite{Shacham:2004:EAR:1030083.1030124,Strackx:2009:BMS:1519144.1519145}.
\\
Additionally, one can make use of stack canaries to try to \emph{detect manipulations} by an attacker. 
This uses a random value, called a \emph{stack canary}. 
This random value is placed in the memory, before any critical information, for example the \emph{return address}.
The return address is a critical piece of information that helps to organize control flow when calling and returning from functions.
Even if an attacker can somehow overwrite the return address in memory, for example using a buffer overflow\cite{OVSPaper}, it is likely that this will change the value of the stack canary as well.
Stack canaries are a detection technique, as the application can check the value of the stack canary against the known random value.
If this value changed, then the attacker's manipulation is detected, and execution can be stopped.
\\[1em]
Other techniques work by providing \emph{isolation} of software components, and introducing security guarantees to memory access using access control.
This way these isolated software components cannot be compromised by attacking other software components that it interacts with.
In other words, isolation of software components compartmentalizes these software components, in a way that compromising is only possible on an individual compartment basis.
This is the goal of secure compilation.

An example of such research can be found in Agten et al.~\cite{Agten:2012:SCM:2354412.2355247}, who already provided a secure compilation scheme for an object based language, when access to memory is restricted based on the value of the program counter. This technique is called \emph{Program Counter Based Access Control}, or \emph{PCBAC}~\cite{PCBAC}.
Later work by Patrignani et al.~\cite{Patrignani} introduced additional object oriented concepts to the fully abstract compilation scheme.

The restricted access of memory can be implemented in hardware~\cite{Sancus,SGX} or using software~\cite{Fides,Salus}.
A recent innovation in restricting access on a hardware level is Intel\textregistered Software Guard Extensions, or \emph{SGX}~\cite{SGX}.

The choice between using software or hardware to provide restricted access to the memory is very important.
For example, this choice affects the size of the trusted computing base or \emph{TCB}.
The TCB of an application contains all components, whether they are hardware or software, on which the application can only place its trust to ensure correct execution.
Even with fully abstract compilation, security issues in the TCB could lead to low-level attacks.
Consequently, a good TCB is a small, and verifiable TCB.

\section{Future Work}
\label{sec:FutureWork}
The \MiniML\ language as described in \myref{chap}{chap:FormalizationOfAdvancedConcepts} still does not provide the full feature set of ML. This section proposes some valuable extensions that could be made to the language.

\begin{itemize}
\item
Currently, the \MiniML\ language allows only a very restricted set of types to be communicated between the secure and insecure code. 
Arrays and pairs can be used within a structure or to implement an opaque type, but they cannot be the argument or return value of a publicly available function or the type of a public field.
This is not a very limiting restriction: a structure can be created to implementing an abstract data type (ADT) that behaves like a list.

However, it could still be a valuable expansion to the language to once again allow lists and pairs as a basic and first class type in the Module language.
As mentioned in \myref{sec}{sec:insecurevaluessecureclosure}, the call-by-value~\cite{Milner:1997:DSM:549659} semantics and declarative style of the ML source language do make this addition more difficult.

For example, when passing a list from the insecure context to a function in the secure code, in source-level semantics this list is passed as a value, and its content is immutable.

As the insecure memory is readable by secure code, the target language translation of the function can read the list directly from the memory, without passing execution to the insecure code.
As the attack described in \myref{sec}{sec:insecurevaluessecureclosure} shows, this capability can introduce security risks if the inherently mutable  memory is changed during a callback to insecure code.
\item
The full-fledged ML language is aware of mutable memory locations, using the concept of a reference type \lsttext{ref $\ \tau$}.
Future work could add this type to \MiniML.

\item
\MiniML\ functors are monadic, they only take a single argument structure.
There are several ways of introducing polyadic functors into \MiniML.
One would work with the current version of \MiniML:
A functor taking multiple arguments could be decomposed in several functors that take a single argument, and output a structure for the next functor to be applied to. 
A drawback to this is that the current \MiniML\ implementation would make each of the intermediate structures available.

Possibly nicer ways of implementing this functionality are:
\begin{itemize}
\item Implementing polyadic functors as real functors whose stubs take more than one frame argument.
Additional difficulties would represent themselves in the way structures are represented by frames.
\item The addition of substructures to the module language.
\end{itemize}
\item
As \LLVMIR\ is not architecture aware, not all necessary security precautions can be described in \LLVMIR.
Specifically, methods of manipulating the stack pointer or the return address are either experimental or nonexistent.

These security precautions are mentioned here, and should be added when the \LLVMIR\ is eventually compiled to machine code for a \emph{Protected Module Platform}~\cite{PCBAC} by the LLVM backend.
\end{itemize}


\section{Conclusion}
\label{sec:Conclusion}
The goal of this thesis was to describe a secure compilation scheme for a language that implements ML-style modules.
\begin{itemize}
\item
\myref{chap}{chap:ACompilationExample} and \myref{chap}{chap:formalspecification} of this thesis describe and formalize a basic version of \MiniML\ and its secure compilation scheme.
Many of the concepts introduced in literature regarding the secure compilation other, object oriented languages can be reused with only small modifications.

\item
\myref{chap}{chap:AdvancedConcepts} expanded the \MiniML\ language, with \myref{chap}{chap:FormalizationOfAdvancedConcepts} providing a formalization of the additions.
While ML is  still a larger and more capable language than the \MiniML\ language of \myref{chap}{chap:FormalizationOfAdvancedConcepts}, the final version of the language does include functors, one of the most important features of ML's powerful module language, as well as closures.
\end{itemize}

\myref{sec}{sec:formalizedcompiler2} formalized a compilation scheme for these additions, bringing secure compilation to an ML-style module language, thereby achieving the original goal of this thesis. 
%Additionally, an informal proof was given in \myref{chap}{chap:InformalProof} that suggests that this compilation is indeed a secure compilation.